Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-03-03 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex
Added estimate on bias term of kernel smoothing when d = 1.

\subsection{Spectral series methods for Sobolev Spaces}
Spectral series estimators are particularly designed for estimation in Sobolev spaces. Let $f_1,f_2,\ldots$ be the unit norm eigenfunctions of $\Delta_P$. The classical order-$K$ spectral series estimator $\wt{f}$ and test statistic $\wt{T}$ are
\begin{equation*}
\wt{f} = \sum_{k = 1}^{K} \dotp{Y}{f_k}_n f_k~~\textrm{and}~~\wt{T} = \|\wt{f}\|_P^2.
\end{equation*}
The estimator $\wt{f}$ and a test based on $\wt{T}$ are minimax rate-optimal over $H_0^s(\mc{X})$, a fact intimately related to the series representation of periodic Sobolev spaces, e.g.~\eqref{eqn:periodic_sobolev_space}.

@@

\subsection{First-order Sobolev classes}
\label{sec:first_order_sobolev_classes}
To begin, we handle the case where $f_0 \in H^1(P)$. We show that $\wh{f}$ and a test based on $\wh{T}$ are minimax optimal, for all values of $d$, and under no additional assumptions on the data generating process, i.e. on $P$ or $f_0$.

@@

\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	The graph radius $\varepsilon$ and number of eigenvectors $K$ satisfies
	\begin{equation*}
	C_0 \cdot \max\biggl\{\biggl(\frac{\log}{n}\biggr)^{1/d}, n^{-1/(2r + d)}\biggr\} \leq \varepsilon \leq \min\{i_0, K^{-1/d}\}
	\end{equation*}
	and
	\begin{equation*}
	K \asymp (M^2n)^{2d/(2s + d)}
	\end{equation*}
\end{enumerate}

\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_ho}
	Under appropriate conditions, with probability at least $1 - \delta$,
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_ho}
	\|\wh{f} - f_0\|_n^2 \lesssim n^{-2s/(2s + d)}
	\end{equation}
\end{theorem}
Theorem~\ref{thm:laplacian_eigenmaps_estimation_ho}, in combination with Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo}, implies that in the flat Euclidean setting Laplacian eigenmaps is a rate-optimal estimator over Sobolev classes, for all scalings of $s$ and $d$. Some other remarks:
\begin{itemize}
	\item \textbf{(1.1) Optimal rates, no RKHS.} It is worth pointing out that we do not require that the regularity of the Sobolev space satisfy $2s > d$, a condition often seen in the literature. In the ``sub-critical'' regime $2s < d$, the Sobolev space $H^s(\mc{X})$ is quite irregular. For instance, it is not a Reproducing Kernel Hilbert Space, nor does it embed into any H\"{o}lder space. As a result, certain versions of the nonparametric regression problem are ill-posed---for instance when loss is measured in $\Leb^{\infty}$ norm, or when the design points $X$ are assumed to be fixed. Likewise, certain estimators are ``off the table'', most notably RKHS-based methods such as smoothing splines. However, at these scales spectral series estimators still obtain the optimal rates for random design regression measured in $\Leb^2(P)$ norm. Theorem~\ref{thm:laplacian_eigenmaps_estimation_ho} shows that the same is true with respect to Laplacian eigenmaps, with error measured in $\Leb^2(P_n)$ norm.
	\item \textbf{(1.2) The zero-trace condition.} The zero-trace condition can likely be weakened. However, requiring some boundary condition on $f_0$ and its first through $(k -1 )$th-order derivatives is unavoidable. \textcolor{red}{Explain why: graphs are implicitly boundaryless objects, and the eigenvectors of $L$ converge to those eigenvectors of $\Delta_P$ which satisfy boundary conditions. Repeat that this is a standard assumption made when analyzing spectral series estimators}.
	\item \textbf{(1.3) The bounded norm and smooth density conditions.} We require that $f_0$ be bounded in the Sobolev norm $\|f_0\|_{H^s(\mc{X})}$---as opposed to the semi-norm $|f_0|_{H^s(\mc{X})}$--- and that $p \in C^{s - 1}(\mc{X})$. Both requirements are related and fundamental, as we shall now explain. As previously mentioned, $L$ converges to a weighted-version of the Laplace operator $\Delta_P$ (also known as the Fokker-Planck) operator. The operator $\Delta_P$ includes both first and second-order derivatives,
	\begin{equation}
	\label{eqn:fokker_planck_1}
	\Delta_Pf= \frac{-1}{p} \mathrm{div}(p^2 \nabla f) = \Delta f - 2\frac{\nabla p^{\top} \nabla f}{p},
	\end{equation}
	and therefore a function $f \in H_0^s(\mc{X})$ has finite energy $\|\Delta_P f\|_{P}^2$ only if the first-order partial derivatives of $f$ and $p$ are bounded. \textcolor{red}{This discussion---and in particular the last sentence---are obviously not quite correct. This is in part because you realize now that the requirements are not in fact fundamental. Instead, the requirement should merely be that $\|\Delta_P^kf\|_{\Leb^2(P)}^2$ is bounded---here $k = s/2$ and we assume $s$ is even---and the upper bounds should be in terms of this semi-norm.}.
	\item \textbf{(1.4) Both of the remarks after Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} continue to apply.}
\end{itemize}

\textbf{(1.4) Rates for testing with higher-order smoothness assumptions.} For testing, we can get the optimal rate $n^{-4s/(4s + d)}$ when $d \leq 4$. When $4 < d \leq 4s$, we get the rate~\textcolor{red}{(...)} Recall that the rate $n^{-4s/(4s + d)}$ is optimal whenever $d \leq 4s$. Therefore there are scalings of $s$ and $d$ for which we do not get the optimal rate. We do not know whether this is due to looseness in our proof techniques, or a problem with the test itself, and leave this for future work.

@@

\textbf{(2) Upper bound on the semi-norm. } Proposition~\ref{prop:graph_seminorm} gives a sufficient upper bound on $\dotp{L^s f_0}{f_0}_n$. Take $\beta := s - 1$. 
\begin{proposition}
	\label{prop:graph_seminorm} 
	Suppose $f \in H_0^s(\mc{X})$, that $p \in C^{\beta}(\mc{X})$, and that $r \gtrsim n^{-1/d}$. Then with probability at least $1 - \delta$, it holds that 
	\begin{equation}
	\label{eqn:graph_seminorm_1}
	\dotp{L^s f}{f}_n \leq \frac{C}{\delta} \|f\|_{H^s} \|p\|_{C^{\beta}} + \textcolor{red}{(?)}
	\end{equation}
	Therefore, if additionally $r \gtrsim n^{-1/(2\beta + d)}$, then 
	\begin{equation}
	\label{eqn:graph_seminorm_2}
	\dotp{L^s f}{f}_n \leq \frac{C}{\delta} \|f\|_{H^s} \|p\|_{C^{\beta}}
	\end{equation}
	When $s = 1$, the results hold for $f \in H^1(\mc{X})$ and any density $p$ that satisfies Assumption~\ref{asmp:density}.
\end{proposition}
The proof of Proposition~\ref{prop:graph_seminorm} involves two steps. In the first step, we show that with probability at least $1 - \delta$, the graph semi-norm $\dotp{L^s f}{f}_n$ is upper bounded by a non-local continuum energy $E_{\varepsilon}^{(s)}(f)$, defined as
\begin{equation*}
E_{\varepsilon}^{(s)}(f) := \dotp{L_{P,\varepsilon}^sf}{f}_{P}
\end{equation*}
where $L_{P,\varepsilon}$ is a non-local approximation to $\Delta_P$, 
\begin{equation*}
L_{P,\varepsilon}f(x) := \int_{\mc{X}}\bigl(f(z) - f(x)\bigr) \eta\biggl(\frac{\|z - x\|}{\varepsilon}\biggr) \,dP(x).
\end{equation*}
In the second step, we show that $E_{\varepsilon}^{(s)}(f)$ is in turn upper bounded by a constant times $D_s(f) := \|f\|_{H^s(P)}^2$. Together, these steps establish Proposition~\ref{prop:graph_seminorm}.

When $s = 1$, similar results have been shown in \textcolor{red}{(?)}. On the other hand, much less work has been done in the $s > 1$ case, where the analysis is more complicated. For one thing, we note that when $s > 1$ the graph energy $\dotp{L^s f}{f}_n$ is itself a biased estimate of the non-local energy $E_{\varepsilon}^{(s)}(f)$. This bias term is the second term on the right hand side of~\eqref{eqn:graph_seminorm_1}, and is neglible only when $r \gtrsim n^{-1/(2r + d)}$. On the other hand, in order to give an estimate of $E_{\varepsilon}^{(s)}(f)$ in terms of the Sobolev norm $D_s(f)$, we first show that $L_{P,\varepsilon}^jf(x) \approx \Delta_P^jf(x)$, for $j = s/2$ when $s$ is odd and $j = (s - 1)/2$ when $s$ is even. In words, we show that a $j$th-order iterated difference operator approximates a $2j$-th (or $2j + 1$-th) order differential operator.  It is in order to establish this fact that we require on the one hand $p \in C^{r}(\mc{X};M)$, and on the other hand that $f$ be zero-trace.

@@

\paragraph{Testing.}

@@

Of course, the Laplacian eigenmaps estimator is only defined at the observed design points $X_1,\ldots,X_n$, and to measure its error in $L^2(P)$ norm we must first give a means of extending it to the rest of $\Xset$. We propose a simple method, essentially kernel smoothing, to do the job, which can be applied to any estimator defined at the design points, including Laplacian eigenmaps. We show that the smoothed version of $\wh{f}$ has optimal $L^2(P)$ error.

>>>>

\subsection{Spectral series methods for Sobolev Spaces}
Spectral series estimators---and tests using their $\Leb^2$-norm---are particularly designed for regression in Sobolev spaces. Let $f_1,f_2,\ldots$ be the unit norm eigenfunctions of $\Delta_P$. The classical order-$K$ spectral series estimator $\wt{f}$ and test statistic $\wt{T}$ are
\begin{equation*}
\wt{f} = \sum_{k = 1}^{K} \dotp{Y}{f_k}_n f_k~~\textrm{and}~~\wt{T} = \|\wt{f}\|_P^2.
\end{equation*}
The estimator $\wt{f}$ and a test based on $\wt{T}$ are minimax rate-optimal over $H_0^s(\mc{X})$, a fact intimately related to the series representation of periodic Sobolev spaces, e.g.~\eqref{eqn:periodic_sobolev_space}.

@@

\subsection{First-order Sobolev classes}
\label{sec:first_order_sobolev_classes}
To begin, we handle the case where $f_0 \in H^1(P)$. We show that $\wh{f}$ and a test based on $\wh{T}$ are minimax optimal, for all values of $d$, and under no additional assumptions on the data generating process, i.e. on either $P$ or $f_0$.

@@

\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp:parameters_estimation_ho}
	The graph radius $\varepsilon$ and number of eigenvectors $K$ satisfy
	\begin{equation*}
	C_0 \cdot \max\biggl\{\biggl(\frac{\log}{n}\biggr)^{1/d}, n^{-1/(2r + d)}\biggr\} \leq \varepsilon \leq \min\{i_0, K^{-1/d}\}
	\end{equation*}
	and
	\begin{equation*}
	K \asymp (M^2n)^{2d/(2s + d)}
	\end{equation*}
\end{enumerate}

\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_ho}
	Under appropriate conditions, with probability at least $1 - \delta$,
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_ho}
	\|\wh{f} - f_0\|_n^2 \lesssim n^{-2s/(2s + d)}
	\end{equation}
\end{theorem}
Theorem~\ref{thm:laplacian_eigenmaps_estimation_ho}, in combination with Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo}, implies that in the flat Euclidean setting Laplacian eigenmaps is a rate-optimal estimator over Sobolev classes, for all scalings of $s$ and $d$. Some other remarks:
\begin{itemize}
	\item \textbf{(1.1) Optimal rates, no RKHS.} It is worth pointing out that we do not require that the regularity of the Sobolev space satisfy $2s > d$, a condition often seen in the literature. In the ``sub-critical'' regime $2s < d$, the Sobolev space $H^s(\mc{X})$ is quite irregular. For instance, it is not a Reproducing Kernel Hilbert Space, nor does it embed into any H\"{o}lder space. As a result, certain versions of the nonparametric regression problem are ill-posed---for instance when loss is measured in $\Leb^{\infty}$ norm, or when the design points $\{X_1,\ldots,X_n\}$ are assumed to be fixed. Likewise, certain estimators are ``off the table'', most notably RKHS-based methods such as smoothing splines. Nevertheless, for random design regression with error measured in $\Leb^2(P)$-norm, spectral series estimators obtain the minimax rates for all values of $s$ and $d$. Theorem~\ref{thm:laplacian_eigenmaps_estimation_ho} shows that the same is true with respect to Laplacian eigenmaps, with error measured in $\Leb^2(P_n)$ norm.
	\item \textbf{(1.2) The zero-trace condition.} The zero-trace condition can likely be weakened. However, requiring some boundary condition on $f_0$ and its first through $(k -1 )$th-order derivatives is unavoidable. \textcolor{red}{Explain why: graphs are implicitly boundaryless objects, and the eigenvectors of $L$ converge to those eigenvectors of $\Delta_P$ which satisfy boundary conditions. Repeat that this is a standard assumption made when analyzing spectral series estimators}.
	\item \textbf{(1.3) The bounded norm and smooth density conditions.} We require that $f_0$ be bounded in the Sobolev norm $\|f_0\|_{H^s(\mc{X})}$---as opposed to the semi-norm $|f_0|_{H^s(\mc{X})}$--- and that $p \in C^{s - 1}(\mc{X})$. Both requirements are related and fundamental, as we shall now explain. As previously mentioned, $L$ converges to a weighted-version of the Laplace operator $\Delta_P$ (also known as the Fokker-Planck) operator. The operator $\Delta_P$ includes both first and second-order derivatives,
	\begin{equation}
	\label{eqn:fokker_planck_1}
	\Delta_Pf= \frac{-1}{p} \mathrm{div}(p^2 \nabla f) = \Delta f - 2\frac{\nabla p^{\top} \nabla f}{p},
	\end{equation}
	and therefore a function $f \in H_0^s(\mc{X})$ has finite energy $\|\Delta_P f\|_{P}^2$ only if the first-order partial derivatives of $f$ and $p$ are bounded. \textcolor{red}{This discussion---and in particular the last sentence---are obviously not quite correct. This is in part because you realize now that the requirements are not in fact fundamental. Instead, the requirement should merely be that $\|\Delta_P^kf\|_{\Leb^2(P)}^2$ is bounded---here $k = s/2$ and we assume $s$ is even---and the upper bounds should be in terms of this semi-norm.}.
	\item \textbf{(1.4) Both of the remarks after Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} continue to apply.}
\end{itemize}

\textbf{(1.4) Rates for testing with higher-order smoothness assumptions.} For testing, we can get the optimal rate $n^{-4s/(4s + d)}$ when $d \leq 4$. When $4 < d \leq 4s$, we get the rate~\textcolor{red}{(...)} Recall that the rate $n^{-4s/(4s + d)}$ is optimal whenever $d \leq 4s$. Thus when $4s > d$, our upper bounds do not imply that $\varphi$ is an optimal test. We do not know whether this is due to looseness in our proof techniques, or a problem with the test itself, and leave this for future work.

@@

\textbf{(2) Upper bound on the semi-norm. } Proposition~\ref{prop:graph_seminorm} gives a sufficient upper bound on $\dotp{L^s f_0}{f_0}_n$. Take $\beta := s - 1$. 
\begin{proposition}
	\label{prop:graph_seminorm} 
	Suppose $f \in H_0^s(\mc{X})$, that $p \in C^{\beta}(\mc{X})$, and that $r \gtrsim n^{-1/d}$. Then with probability at least $1 - \delta$, it holds that 
	\begin{equation}
	\label{eqn:graph_seminorm_1}
	\dotp{L^s f}{f}_n \leq \frac{C}{\delta} \|f\|_{H^s} \|p\|_{C^{\beta}} + \textcolor{red}{(?)}
	\end{equation}
	Therefore, if additionally $r \gtrsim n^{-1/(2\beta + d)}$, then 
	\begin{equation}
	\label{eqn:graph_seminorm_2}
	\dotp{L^s f}{f}_n \leq \frac{C}{\delta} \|f\|_{H^s} \|p\|_{C^{\beta}}
	\end{equation}
	When $s = 1$, the results hold for $f \in H^1(\mc{X})$ and any density $p$ that satisfies Assumption~\ref{asmp:density}.
\end{proposition}
The proof of Proposition~\ref{prop:graph_seminorm} involves two steps. In the first step, we show that with probability at least $1 - \delta$, the graph semi-norm $\dotp{L^s f}{f}_n$ is upper bounded by a non-local continuum energy $E_{\varepsilon}^{(s)}(f)$, defined as
\begin{equation*}
E_{\varepsilon}^{(s)}(f) := \dotp{L_{P,\varepsilon}^sf}{f}_{P}
\end{equation*}
where $L_{P,\varepsilon}$ is a non-local approximation to $\Delta_P$, 
\begin{equation}
\label{eqn:nonlocal_laplacian}
L_{P,\varepsilon}f(x) := \frac{1}{\varepsilon^{d + 2}}\int_{\mc{X}}\bigl(f(z) - f(x)\bigr) \eta\biggl(\frac{\|z - x\|}{\varepsilon}\biggr) \,dP(x).
\end{equation}
In the second step, we show that $E_{\varepsilon}^{(s)}(f)$ is in turn upper bounded by a constant times $D_s(f) := \|f\|_{H^s(P)}^2$. Together, these steps establish Proposition~\ref{prop:graph_seminorm}.

When $s = 1$, similar results have been shown in \textcolor{red}{(?)}. On the other hand, much less work has been done in the $s > 1$ case, where the analysis is more complicated. For one thing, we note that when $s > 1$ the graph energy $\dotp{L^s f}{f}_n$ is itself a biased estimate of the non-local energy $E_{\varepsilon}^{(s)}(f)$. This bias term is the second term on the right hand side of~\eqref{eqn:graph_seminorm_1}, and is neglible only when $r \gtrsim n^{-1/(2r + d)}$. On the other hand, in order to give an estimate of $E_{\varepsilon}^{(s)}(f)$ in terms of the Sobolev norm $D_s(f)$, we first show that $L_{P,\varepsilon}^jf(x) \approx \Delta_P^jf(x)$, for $j = s/2$ when $s$ is odd and $j = (s - 1)/2$ when $s$ is even. In words, we show that a $j$th-order iterated difference operator approximates a $2j$-th (or $2j + 1$-th) order differential operator.  In to establish this fact, we require that on the one hand $p \in C^{r}(\mc{X};M)$, and on the other hand $f$ be zero-trace.

@@

\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp:parameters_estimation_manifold}
	The graph radius $\varepsilon$ and number of eigenvectors $K$ satisfy
	\begin{equation}
	\label{eqn:radius_estimation_manifold}
	C_0 \cdot \max\biggl\{\biggl(\frac{\log}{n}\biggr)^{1/m}, n^{-1/(2r + m)}\biggr\} \leq \varepsilon \leq \min\{i_0, K^{-1/m}\}
	\end{equation}
	and
	\begin{equation*}
	K \asymp (M^2n)^{2m/(2s + m)}
	\end{equation*}
\end{enumerate}

\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_manifold}
	Suppose $f_0 \in H^s(\mc{X},M)$, for $s \leq 2$. Under appropriate conditions, with probability at least $1 - \delta - \textcolor{red}{(?)}$,
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_manifold}
	\|\wh{f} - f_0\|_n^2 \lesssim M^{2m/(2s + m)}n^{-2s/(2s + m)}
	\end{equation}
\end{theorem}

\paragraph{Testing.}
To construct a minimax optimal test using $\wh{T}$, we choose $\varepsilon$ and $K$ as in~\ref{asmp:parameters_testing_fo}, except with the ambient dimension $d$ replaced by the intrinsic dimension $m$.
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp:parameters_testing_manifold}
	The graph radius $\varepsilon$ and number of eigenvectors $K$ satisfy~\eqref{eqn:radius_estimation_manifold}, for $r = 0$. Additionally, there exist constants $c_2$ and $C_2$ such that
	\begin{equation*}
	c_2 (M^2n)^{2m/(4s + m)} \leq K \leq C_2 (M^2n)^{2m/(4s + m)}
	\end{equation*}
\end{enumerate}

\begin{theorem}
	\label{thm:laplacian_eigenmaps_testing_manifold}
	Suppose $f_0 \in H^1(\mc{X};M)$. If the Laplacian eigenmaps test $\varphi$ is computed with parameters $\varepsilon$ and $K$ that satisfy~\ref{asmp:parameters_testing_manifold}, then the following statement holds for any $\delta \in (0,1)$: if $f_0$ satisfies
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_testing_criticalradius_manifold}
	\|f_0\|_P^2 \geq \frac{C}{\delta} \max\biggl\{\frac{1}{\delta}M^{2m/(4 + m)}n^{-4/(4 + m)}, M n^{-p(m)}\biggr\},
	\end{equation}
	then $\varphi$ has Type II error upper bounded by 
	\begin{equation}
	\Pbb_{f_0}(\varphi = 0) \leq \delta + \textcolor{red}{(...)}
	\end{equation}
\end{theorem}

\begin{itemize}
	\item \textbf{Proof strategy.} The proofs of Theorems~\ref{thm:laplacian_eigenmaps_estimation_manifold} and~\ref{thm:laplacian_eigenmaps_testing_manifold} proceed in a similar manner to the full-dimensional setting.
	\item \textbf{Restriction on number of derivatives.} Unlike in the full-dimensional case, Theorem~\ref{thm:laplacian_eigenmaps_estimation_manifold} holds only when $s \leq 2$. Roughly speaking, we need this restriction in order to show that $L_{P,\varepsilon}^sf(x)$ is a good approximation to $\Delta_P^sf(x)$. Otherwise, the error incurred by using Euclidean distance rather than geodesic distance in~\eqref{eqn:nonlocal_laplacian} will dominate.
\end{itemize}

@@

Of course, the Laplacian eigenmaps estimator is only defined at the observed design points $X_1,\ldots,X_n$, and to measure its error in $L^2(P)$ norm we must first extend it to the rest of $\Xset$. We propose a simple method, essentially kernel smoothing, to do the job, which can be applied to any estimator defined at the design points, including Laplacian eigenmaps. We show that the smoothed version of $\wh{f}$ has optimal $L^2(P)$ error.
